### 2023-05-04 submission/manuscript.Rmd
First full draft of Methods

```{r}
gp_median <- as.integer(median(gp_counts$n_seqs))
gp_mean <- format(round(mean(gp_counts$n_seqs), digits = 1), nsmall = 1)
gp_sd <- sd(gp_counts$n_seqs)
gp_min <- as.integer(min(gp_counts$n_seqs))
gp_max <- as.integer(max(gp_counts$n_seqs))
gp_max_min_ratio <- format(round(gp_max/gp_min, digits = 1), nsmall = 1)

@@

I am grateful to McMurdie and Holmes for providing the source code that they used to conduct their analysis as R-flavored markdown files. This provided me with a better understanding of their methods including noticing that only a single subsampling step was performed for each random seed and that only 5 random seeds were used. Furthermore, their code enabled me to replicate and build upon their simulations. I have used alternative simulation and evaluation strategies to look closer at rarefaction and its alternatives and the practice of filtering low abundance seqeunces. The results of those studies are similar to this reanalysis. Rarefaction of unfiltered datasets yields the most robust results.

@@

@@

>>>>

```{r}
gp_median <- as.integer(median(gp_counts$n_seqs))
gp_mean <- format(round(mean(gp_counts$n_seqs), digits = 1), nsmall = 1)
gp_sd <- sd(gp_counts$n_seqs)
gp_min <- as.integer(min(gp_counts$n_seqs))
gp_max <- as.integer(max(gp_counts$n_seqs))
gp_max_min_ratio <- format(round(gp_max / gp_min, digits = 1), nsmall = 1)

@@

In a parallel set of analyses I have used alternative simulation and evaluation strategies to look closer at rarefaction and its alternatives and the practice of filtering low abundance seqeunces. The results of those studies are similar to this reanalysis. These studies demonstrate that sequence and OTU filtering "requires the omission of available valid data" and that rarefaction is the only available method for controlling the effects of uneven sampling. Far from being inadmissible, rarefaction of unfiltered datasets yields the most robust results.

@@

### Code availability

A git repository containing all of the code needed to reproduce this study is available at https://www.github.com/SchlossLab/Schloss_WNWN_MSpectrum_2023. All simulations and analyses were performed using R and bash scripts using Snakemake (v7.24.0) to track dependencies and automate the pipeline and conda (v4.12.0) and mamba (v1.1.0) to specify software and package versions (see workflow/envs/nr-base.yml in repository). This manuscript was written as an R-flavored markdown document and rendered using the `rmarkdown` R package (v2.18). All figures were generated using `dplyr` (v1.1.0) and `ggplot2` (v3.4.2) from the `tidyverse` metapackage (v1.3.2) and `ggtext` (v0.1.2) and `ggh4x` (v0.2.4) within R (v.4.2.2; see workflow/envs/nr-modern.yml in repository).

### Reproducing WNWN Protocol S1

The compressed directory published with WNWN as Protocol S1 is available on the PLOS Computational Biology website with the original paper at https://doi.org/10.1371/journal.pcbi.1003531.s001. To render the R-flavored markdown files to HTML files using software and packages as similar as possible to that of the original study, I created an environment (see workflow/envs/nr-s1.yml in repository) that contained versions as close to those indicated in the pre-rendered files. The most important packages for the reproduction of the cluster analysis included `DESeq` (my version: v1.39.0 vs. WNWN version: v1.14.0), `edgeR` (v3.30.0 vs. v3.4.0), `cluster` (v2.1.0 vs v1.14.4), `phyloseq` (v1.32.0 vs. v1.6.0), and `PoiClaClu` (v1.0.2.1 vs. 1.0.2). In addition, I used R v4.0.2 whereas the original WNWN analysis used v3.0.2. Finally, to render the `simulation-cluster-accuracy-server.Rmd` file to HTML, it was necessary to apply a patch to the code. This patch set the path to the location where the figures should be saved and removed the R code that deleted all objects in the enviornment. Both changes were necessary to better organize the project and had no bearing on the content of the HTML file that was rendered.

### Simulated communities

The code contained within `simulation-cluster-accuracy-server.Rmd` was spilt into individual R scripts and modified to make the generation of the simulated communities more modular and scalable. The code for geneating the simulated communities was run using the nr-s1 conda environment. The Ocean and Feces parent distributions were generated as described above. Five variables were altered to simulate each set of communities. First, eight mixing fractions were used to manipulate the effect size between the two simulated treatment groups with 40 samples per group. These were the same values used in the original WNWN study (1.0 [i.e., no difference], 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5). Second, the number of sequences in each of the 80 samples was either randomly selected from the `r gp_n_samples` library sizes contained wihtin the GlobalPatterns dataset or from a log distribution of 80 sequencing depths evenly distributed between the smallest (`r gp_min` sequences) and largest dataset (`r gp_max` sequences) in the GlobalPatterns dataset (Figure **distribution_shape.pdf**). Third, the resulting 80 sequencing depths were scaled so that the median sequencing depth across the 80 samples was either 1,000, 2,000, 5,000, 10,000, and 50,000 sequences. Fourth, the sequencing depth of each sample was either randomly assigned to each treatment group or assigned so that those depths less than or greater than the median were assigned to separate treatment groups. Finally, each set of conditions was replicated 100 times. These five parameters resulted in 1,600 simulated datasets (8 fractions x 2 sequencing depth models x 5 sequencing depths x 2 sample assignment models x 100 replicates).

### Generation of distances between communities

Next, each simulation was further processed by filtering rare OTUs, to normalize uneven sampling depths, and to calculate ecological distances between samples. Again, the code contained within `simulation-cluster-accuracy-server.Rmd` was spilt into individual R scripts and modified to make the generation of the simulated communities more modular and scalable within the nr-s1 conda environment. First, each simulation was either filtered to remove rare OTUs or left unfilterd. As was done in the original WNWN study, filtering was performed in two steps: (1) any OTUs whose total abundance was less than 3 across all 80 samples and that did not appear in at least 3 samples and (2) any OTU that did not have more than 1 sequence in more than 5% of the 80 samples (i.e., 4 samples) and that did not have a total abundance across the 80 samples greater than one half of the number of samples in each community type (i.e., 20) were removed. Second, OTU data was normalized by one of six approaches: (1) non-normalized raw counts, (2) relative abundance, (3) varaiance stabilization using DESeq, (4) Upper Quartile normalization using edgeR, (5) a single subsampling to a common number of sequences, (6) rarefaction with 100 randomizatoins to a common number of sequences. Although the data were not presented, the original and my code also normalized the counts using trimmed mean of M-values (TMM) and relative log expression (RLE) in edgeR. The level of subsampling and rarefaction was selected based on the size of the sample whose sequencing depth was at the 0 (i.e., included all of the samples), 5, 10, 15, 20, 25, and 40th percentiles. Finally, the normalized simulated datasets were used to calculate pairwise distances between the 80 samples using seven different calculations: Bray-Curtis, Euclidean, Poisson, Mean Squared Difference, Unweighted UniFrac, Weighted Unifrac, and Biological Coeffcient of Variation (BCV); however, data using BCV were not discussed in this or the original analysis. For the rarefaction normalized data, each of the 100 subsampled datasets were used to calculate a distance matrix. The mean of the 100 distance matrices was used as the distance matrix for the rarefaction data. As indicated in the Results section, there were combinations of normalization and distance methods that were not compatible (e.g., DESeq Variance Stabilization normalization with Bray-Curtis distances). In such cases, distance matrices were coded as `NA`. My choice of combinations of normalziations and distances to include was determined by the original analysis except where noted. For each simulated dataset there were 280 possible combinations of filtering, normalization, and distance methods (2 filtering methods x 20 normalization method x 7 distance methods).

### Analysis of distances between communities

Two general strategies were used to analyze the pairwise distances in each distance matrix. First, by recycling the the code contained within `simulation-cluster-accuracy-server.Rmd` in the nr-s1 conda environment, samples were assigned to one of two clusters using partitioning around the medoid (PAM) with the `pam` function from the `cluster` package, K-means clustering with the `kmeans` function from the `stats` base package, and hierarchical clustering with the `hclust` and `cutree` functions from the `stats` base package. The accuracy of the clustering was measured as the fraction of the 80 samples assigned to the correct cluster. As in the original analysis, if samples were removed because the number of sequences in them fell below the threshold for subsamplnig or rarefaction, then the accuracy could be below 50%. For the 1 mixing fraction all of the samples should have been assigned to one group; however, because they were forced into two groups there would be 41 "correctly" assigned samples of the 80 (`r 100 * 41/80`%). Second, the distance matrices were analyzed for significant different centroids using the `adonis2` function from the `vegan` package (v2.6-4) in the nr-modern environment. The fraction of significant tests was the fraction of the 100 replicate simulations that had a p-value less than or equal to 0.05. The false positive rate for a condition was the fraction of significant tests when the mixing level was 1. The power was the fraction of significant tests at other mixing levels.

### Analysis of richness and diversity between communities

The original WNWN analysis did not address the impact of normalization on alpha diversitiy metrics. Using the nr-s1 conda environment, I measured the richness and Shannon diversity using the normalized OTU counts. Richness was measured by counting the number of OTUs for each sample across the simulations and the Shannon diversity using their relative abundance using the commonly used forumla. For the rarefaction normalized data, each of the 100 subsampled datasets were used to calculate the richness and diversity. The mean of the 100 values was used as the value for the rarefaction data. To assess differences between the two treatment groups, the richness and diversity values were compared using the two-sample Wilcoxon non-parametric test with the `wilcox.test` from the `stats` base package. Similar to the adonis analysis, the fraction of significant tests was the fraction of the 100 replicate simulations that had a p-value less than or equal to 0.05. The false positive rate for a condition was the fraction of significant tests when the mixing level was 1. The power was the fraction of significant tests at other mixing levels.

@@

I am grateful to Paul McMurdie and Susan Holmes for publishing the source code that they used to conduct their analysis as R-flavored markdown files in the supplement to WNWN. This provided me with a better understanding of their methods including noticing that only a single subsampling step was performed for each random seed and that only 5 random seeds were used. Furthermore, their code enabled me to replicate and build upon their simulations.

### 2023-05-04 submission/manuscript.tex
First full draft of Methods

I am grateful to McMurdie and Holmes for providing the source code that
they used to conduct their analysis as R-flavored markdown files. This
provided me with a better understanding of their methods including
noticing that only a single subsampling step was performed for each
random seed and that only 5 random seeds were used. Furthermore, their
code enabled me to replicate and build upon their simulations. I have
used alternative simulation and evaluation strategies to look closer at
rarefaction and its alternatives and the practice of filtering low
abundance seqeunces. The results of those studies are similar to this
reanalysis. Rarefaction of unfiltered datasets yields the most robust
results.

@@

>>>>

In a parallel set of analyses I have used alternative simulation and
evaluation strategies to look closer at rarefaction and its alternatives
and the practice of filtering low abundance seqeunces. The results of
those studies are similar to this reanalysis. These studies demonstrate
that sequence and OTU filtering ``requires the omission of available
valid data'' and that rarefaction is the only available method for
controlling the effects of uneven sampling. Far from being inadmissible,
rarefaction of unfiltered datasets yields the most robust results.

@@

\hypertarget{code-availability}{%
\subsubsection{Code availability}\label{code-availability}}

A git repository containing all of the code needed to reproduce this
study is available at
\url{https://www.github.com/SchlossLab/Schloss_WNWN_MSpectrum_2023}. All
simulations and analyses were performed using R and bash scripts using
Snakemake (v7.24.0) to track dependencies and automate the pipeline and
conda (v4.12.0) and mamba (v1.1.0) to specify software and package
versions (see workflow/envs/nr-base.yml in repository). This manuscript
was written as an R-flavored markdown document and rendered using the
\texttt{rmarkdown} R package (v2.18). All figures were generated using
\texttt{dplyr} (v1.1.0) and \texttt{ggplot2} (v3.4.2) from the
\texttt{tidyverse} metapackage (v1.3.2) and \texttt{ggtext} (v0.1.2) and
\texttt{ggh4x} (v0.2.4) within R (v.4.2.2; see
workflow/envs/nr-modern.yml in repository).

\hypertarget{reproducing-wnwn-protocol-s1}{%
\subsubsection{Reproducing WNWN Protocol
S1}\label{reproducing-wnwn-protocol-s1}}

The compressed directory published with WNWN as Protocol S1 is available
on the PLOS Computational Biology website with the original paper at
\url{https://doi.org/10.1371/journal.pcbi.1003531.s001}. To render the
R-flavored markdown files to HTML files using software and packages as
similar as possible to that of the original study, I created an
environment (see workflow/envs/nr-s1.yml in repository) that contained
versions as close to those indicated in the pre-rendered files. The most
important packages for the reproduction of the cluster analysis included
\texttt{DESeq} (my version: v1.39.0 vs.~WNWN version: v1.14.0),
\texttt{edgeR} (v3.30.0 vs.~v3.4.0), \texttt{cluster} (v2.1.0 vs
v1.14.4), \texttt{phyloseq} (v1.32.0 vs.~v1.6.0), and \texttt{PoiClaClu}
(v1.0.2.1 vs.~1.0.2). In addition, I used R v4.0.2 whereas the original
WNWN analysis used v3.0.2. Finally, to render the
\texttt{simulation-cluster-accuracy-server.Rmd} file to HTML, it was
necessary to apply a patch to the code. This patch set the path to the
location where the figures should be saved and removed the R code that
deleted all objects in the enviornment. Both changes were necessary to
better organize the project and had no bearing on the content of the
HTML file that was rendered.

\hypertarget{simulated-communities}{%
\subsubsection{Simulated communities}\label{simulated-communities}}

The code contained within
\texttt{simulation-cluster-accuracy-server.Rmd} was spilt into
individual R scripts and modified to make the generation of the
simulated communities more modular and scalable. The code for geneating
the simulated communities was run using the nr-s1 conda environment. The
Ocean and Feces parent distributions were generated as described above.
Five variables were altered to simulate each set of communities. First,
eight mixing fractions were used to manipulate the effect size between
the two simulated treatment groups with 40 samples per group. These were
the same values used in the original WNWN study (1.0 {[}i.e., no
difference{]}, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5). Second, the
number of sequences in each of the 80 samples was either randomly
selected from the 26 library sizes contained wihtin the GlobalPatterns
dataset or from a log distribution of 80 sequencing depths evenly
distributed between the smallest (58,688 sequences) and largest dataset
(2,357,181 sequences) in the GlobalPatterns dataset (Figure
\textbf{distribution\_shape.pdf}). Third, the resulting 80 sequencing
depths were scaled so that the median sequencing depth across the 80
samples was either 1,000, 2,000, 5,000, 10,000, and 50,000 sequences.
Fourth, the sequencing depth of each sample was either randomly assigned
to each treatment group or assigned so that those depths less than or
greater than the median were assigned to separate treatment groups.
Finally, each set of conditions was replicated 100 times. These five
parameters resulted in 1,600 simulated datasets (8 fractions x 2
sequencing depth models x 5 sequencing depths x 2 sample assignment
models x 100 replicates).

\hypertarget{generation-of-distances-between-communities}{%
\subsubsection{Generation of distances between
communities}\label{generation-of-distances-between-communities}}

Next, each simulation was further processed by filtering rare OTUs, to
normalize uneven sampling depths, and to calculate ecological distances
between samples. Again, the code contained within
\texttt{simulation-cluster-accuracy-server.Rmd} was spilt into
individual R scripts and modified to make the generation of the
simulated communities more modular and scalable within the nr-s1 conda
environment. First, each simulation was either filtered to remove rare
OTUs or left unfilterd. As was done in the original WNWN study,
filtering was performed in two steps: (1) any OTUs whose total abundance
was less than 3 across all 80 samples and that did not appear in at
least 3 samples and (2) any OTU that did not have more than 1 sequence
in more than 5\% of the 80 samples (i.e., 4 samples) and that did not
have a total abundance across the 80 samples greater than one half of
the number of samples in each community type (i.e., 20) were removed.
Second, OTU data was normalized by one of six approaches: (1)
non-normalized raw counts, (2) relative abundance, (3) varaiance
stabilization using DESeq, (4) Upper Quartile normalization using edgeR,
(5) a single subsampling to a common number of sequences, (6)
rarefaction with 100 randomizatoins to a common number of sequences.
Although the data were not presented, the original and my code also
normalized the counts using trimmed mean of M-values (TMM) and relative
log expression (RLE) in edgeR. The level of subsampling and rarefaction
was selected based on the size of the sample whose sequencing depth was
at the 0 (i.e., included all of the samples), 5, 10, 15, 20, 25, and
40th percentiles. Finally, the normalized simulated datasets were used
to calculate pairwise distances between the 80 samples using seven
different calculations: Bray-Curtis, Euclidean, Poisson, Mean Squared
Difference, Unweighted UniFrac, Weighted Unifrac, and Biological
Coeffcient of Variation (BCV); however, data using BCV were not
discussed in this or the original analysis. For the rarefaction
normalized data, each of the 100 subsampled datasets were used to
calculate a distance matrix. The mean of the 100 distance matrices was
used as the distance matrix for the rarefaction data. As indicated in
the Results section, there were combinations of normalization and
distance methods that were not compatible (e.g., DESeq Variance
Stabilization normalization with Bray-Curtis distances). In such cases,
distance matrices were coded as \texttt{NA}. My choice of combinations
of normalziations and distances to include was determined by the
original analysis except where noted. For each simulated dataset there
were 280 possible combinations of filtering, normalization, and distance
methods (2 filtering methods x 20 normalization method x 7 distance
methods).

\hypertarget{analysis-of-distances-between-communities}{%
\subsubsection{Analysis of distances between
communities}\label{analysis-of-distances-between-communities}}

Two general strategies were used to analyze the pairwise distances in
each distance matrix. First, by recycling the the code contained within
\texttt{simulation-cluster-accuracy-server.Rmd} in the nr-s1 conda
environment, samples were assigned to one of two clusters using
partitioning around the medoid (PAM) with the \texttt{pam} function from
the \texttt{cluster} package, K-means clustering with the
\texttt{kmeans} function from the \texttt{stats} base package, and
hierarchical clustering with the \texttt{hclust} and \texttt{cutree}
functions from the \texttt{stats} base package. The accuracy of the
clustering was measured as the fraction of the 80 samples assigned to
the correct cluster. As in the original analysis, if samples were
removed because the number of sequences in them fell below the threshold
for subsamplnig or rarefaction, then the accuracy could be below 50\%.
For the 1 mixing fraction all of the samples should have been assigned
to one group; however, because they were forced into two groups there
would be 41 ``correctly'' assigned samples of the 80 (51.25\%). Second,
the distance matrices were analyzed for significant different centroids
using the \texttt{adonis2} function from the \texttt{vegan} package
(v2.6-4) in the nr-modern environment. The fraction of significant tests
was the fraction of the 100 replicate simulations that had a p-value
less than or equal to 0.05. The false positive rate for a condition was
the fraction of significant tests when the mixing level was 1. The power
was the fraction of significant tests at other mixing levels.

\hypertarget{analysis-of-richness-and-diversity-between-communities}{%
\subsubsection{Analysis of richness and diversity between
communities}\label{analysis-of-richness-and-diversity-between-communities}}

The original WNWN analysis did not address the impact of normalization
on alpha diversitiy metrics. Using the nr-s1 conda environment, I
measured the richness and Shannon diversity using the normalized OTU
counts. Richness was measured by counting the number of OTUs for each
sample across the simulations and the Shannon diversity using their
relative abundance using the commonly used forumla. For the rarefaction
normalized data, each of the 100 subsampled datasets were used to
calculate the richness and diversity. The mean of the 100 values was
used as the value for the rarefaction data. To assess differences
between the two treatment groups, the richness and diversity values were
compared using the two-sample Wilcoxon non-parametric test with the
\texttt{wilcox.test} from the \texttt{stats} base package. Similar to
the adonis analysis, the fraction of significant tests was the fraction
of the 100 replicate simulations that had a p-value less than or equal
to 0.05. The false positive rate for a condition was the fraction of
significant tests when the mixing level was 1. The power was the
fraction of significant tests at other mixing levels.
